Supp mat: time series paper

### 2017-08-14 writing/manuscript_draft.tex
added stacked barplot for density plot

\usepackage{float}
\usepackage{setspace}\doublespacing \usepackage{lineno}
\usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

@@

Long-term census data are necessary to better understand ecological
processes and make management decisions. However, census data are often
expensive, requiring a lot of people-hours and equipment. Little work
has addressed the question, when is a time series of census data long
enough to address a question of interest? Here, we explore two
approaches to address this question: simulations and an empirical
approach. We specifically determine the minimum time series length
required to estimate significant increases or decreases in population
abundance. Importantly, we examine the ability to detect a trend with a
set level of statistical power, which is often neglected in ecological
time series analyses. Using a simulation approach, we demonstrate how
the minimum time series length required increases with weaker trends in
abundance and with higher population variability. In addition, we
examine 868 populations of vertebrate species to determine the minimum
time required to detect changes in their abundance. We show that 10-15
years of continuous monitoring are required in order to achieve a high
level of statistical power. Similar to simulation-approaches, the
minimum time required for field census data strongly depends on trend
strength, population variability, and temporal autocorrelation. These
results point to the importance of sampling populations over long
periods of time. Most studies, especially those less than 10-15 years,
are probably under-powered and potentially misleading.

@@

Naturally, one might ask, how long of a time series is actually
necessary? This question has important implications for both research
and management. Scientifically, we need to know the time series length
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. The IUCN Red List Categories and Criteria suggest under Criterion
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or 3 generations (IUCN 2012). For both scientific
and management questions, because sampling is typically expensive, we do
not want to sample for longer than is necessary (Gerber, DeMaster, and
Kareiva 1999).

@@

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized a model specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organism, 15 years was adequate to assess long-term trends
in abundance. However, these studies, and other past work, have
typically focused either on theoretical aspects of monitoring design or
focused on single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to assess long-term changes in abundance via
simple linear regression. We examine 878 population time series to
estimate this minimum time series length required for detecting
population trends.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses (P. C.
Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been described elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend.

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is
greater than 0.8 (the dotted line), that is the minimum time required
(\(T_{min}\)).

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterion A2 to
classify a species as vulnerable (IUCN 2012).

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015) like body size and generation length. All 547 populations are
within the Aves class. In the supplementary material, we explore how the
minimum time required does not strongly correlate with any biological
characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then we
determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true'', time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1 - \beta\), where \(\beta\) is the
probability of a type II error). Lastly, we determine which subsample
length is required to guarantee we have above a certain threshold of
statistical power, which we will assume is 0.8 by convention (Cohen
1992).

@@

We determined the minimum time series length required to address a
particular question of interest. More specifically we asked, what is the
minimum time series length required to determine, via linear regression,
the long-term trend in population abundance? The minimum time series
length required is that which had high enough statistical power. We
investigated this question through two approaches: simulation-based and
empirical analyses. We briefly describe results from the simulation
approach and then discuss our empirical approach.

@@

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest. We then ran simulated time
series of different lengths. From these simulations we determined the
minimum time series length required to achieve a certain level of power
(which we set to 0.8). We were also able to alter the parameter values
(trend strength and population variability) of the population model to
see how this altered \(T_{min}\).

In line with past work, we found the \(T_{min}\) increases (i.e.~more
time is required) with decreases in trend strength and with increases in
variability (citation).

\subsection{Empirical test}\label{empirical-test}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
expected based on theoretical results (cite papers), the minimum time
series length required is strongly correlated with trend strength,
population variance, and autocorrelation (Fig \ref{fig:correlates}).
Using a generalized linear model, with a Poisson error structure, all
three of these explanatory variables were significant and had large
effect sizes (see Fig. SX). Combined, trend strength, population
variance, and autocorrelation account for about 72.6 of the explained
deviance, which is similar to \(R^2\) (Zuur et al. 2009), in minimum
time series length required.

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None of
these explanatory variables were predictive of the minimum time series
length required (Fig. \ref{fig:biological_correlates}). Some differences
if \(T_{min}\) did occur between biological class (Fig.
\ref{fig:class}), but this can be attributed to the amount populations
are known to fluctuate (i.e.~the variability in population size).

@@

We explored two different approaches to estimate the minimum time series
length required to address a particular question of interest. We
specifically asked, what is the minimum time series length required to
determine long-term changes in abundance using linear regression? The
simulation-based approach has been suggested by others, especially in
situations more complicated than that suited for classic power analysis
(Gerrodette 1987; P. C. Johnson et al. 2015; Bolker 2008). Our
simulations support past work that longer time series are needed when
the trend strength (i.e.~rate of increase or decrease) is weak or when
population variability is high (Gerrodette 1987).

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C, Sarah J Barry, Heather M Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@

\section{Figures}\label{figures}

stuff here \ref{fig:empirical_approach_example}

>>>>

\usepackage{float}
\usepackage{setspace}\doublespacing \usepackage{lineno}
\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
\linenumbers

@@

Long-term census data are necessary to better understand ecological
processes and make management decisions. However, census data are often
expensive, requiring a lot of time and resource. Little work has
addressed the question, when is a time series of census data long enough
to address a question of interest? Here, we explore two approaches to
address this question: simulations and an empirical approach. We
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, we examine the ability to detect a trend with a set level
of statistical power, which is often neglected in ecological time series
analyses. Using a simulation approach, we demonstrate how the minimum
time series length required increases with weaker trends in abundance
and with higher population variability. In addition, we examine 878
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. We show that 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. These results
point to the importance of sampling populations over long periods of
time. We stress the need for verifying a study has enough statistical
power to accurately determine long-term population trends. Most studies,
especially those less than 10-15 years, are probably under-powered and
potentially misleading.

  \renewcommand{\thesection}{\arabic{section}}

@@

Naturally, one might ask, how long of a time series is actually
necessary? This question has important implications for both research
and management. Scientifically, we need to know the time series length
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when population trend over time is actually meaningful or
not. The IUCN Red List Categories and Criteria suggest, under Criterion
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or 3 generations (IUCN 2012). For both scientific
and management questions, because sampling is typically expensive, we
also do not want to sample for longer than is necessary (Gerber,
DeMaster, and Kareiva 1999).

@@

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized a model specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organism, 15 years was adequate to assess long-term trends
in abundance. However, these studies, and other past work, have
typically focused either on theoretical aspects of monitoring design or
focused on only a few species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
several questions. In the main manuscript, we estimate the minimum time
series length required to assess long-term changes in abundance via
simple linear regression. We then examine 878 population time series to
estimate this minimum time series length required for detecting
population trends. We expand on the results in the supplmentary material
by examining other questions of interest.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power for a particular question. This
approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of the increase or decrease (i.e.~the rate of
increase).

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here, statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend.

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
greater than 0.8 (the dotted line), that is the minimum time required
(\(T_{min}\)).

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation.

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015), including body size and generation length. All 547 populations
are within the Aves class. In the supplementary material, we explore how
the minimum time required does not strongly correlate with any
biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then, we
determine the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true'', time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1 - \beta\), where \(\beta\) is the
probability of a type II error). Lastly, we determine which subsample
length is required to guarantee we have above a certain threshold of
statistical power, which we will assume is 0.8 by convention (Cohen
1992).

@@

We determined the minimum time series length (\(T_{min}\)) required to
address a particular question of interest. More specifically we asked,
what is the minimum time series length required to determine, via linear
regression, the long-term population trend? Here, the minimum time
series length required had high enough statistical power (greater than
0.8) for a set significance level of 0.05. We investigated this question
through two approaches: simulation-based and empirical analyses. We
briefly describe results from the simulation approach and then discuss
our empirical approach.

@@

We constructed a general population model where the trend (i.e.~slope
coefficient) over time could be a model parameter. We chose a simple
model, but any other population model could be used (Fig.
\ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice
should be tailored to the population of interest. We then ran simulated
time series of different lengths. From these simulations we determined
the minimum time series length required to achieve a certain level of
statistical power. We were also able to alter the parameter values
(trend strength and population variability) of the population model to
see how this altered \(T_{min}\) (Fig. \ref{fig:theoretical_approach}).

In line with past work (Gerrodette 1987), we found the \(T_{min}\)
increases (i.e.~more time is required) with decreases in trend strength
and with increases in variability (Fig. \ref{fig:theoretical_approach}).

\subsection{Empirical approach}\label{empirical-approach-1}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
expected based on our simulation results and those of others (Rhodes and
Jonzen 2011), the minimum time series length required is strongly
correlated with trend strength, population variance, and autocorrelation
(Fig \ref{fig:correlates}). Using a generalized linear model, with a
Poisson error structure, all three of these explanatory variables were
significant and had large effect sizes (see Table S1). Combined, trend
strength, population variance, and autocorrelation account for about
72.6 of the explained deviance, which is similar to \(R^2\) (Zuur et al.
2009), in minimum time series length required.

We also examined a subset of populations where more explanatory
variables were known (e.g.~body mass, longevity, location). None of
these explanatory variables were predictive of the minimum time series
length required (Fig. \ref{fig:biological_correlates}). Some differences
in \(T_{min}\) did occur between biological class (Fig.
\ref{fig:class}), but this can be attributed to the amount populations
are known to fluctuate (i.e.~the variability in population size).

@@

We explored two different approaches to estimate the minimum time series
length required to address a particular question of interest. We
specifically asked, what is the minimum time series length required to
determine long-term population trends using linear regression? This is
one of the simplest questions one could ask of a time series. We would
expect more involved questions (e.g.~detecting chaos) to require a
larger minimum time series length. The simulation-based approach has
been suggested by others, especially in situations more complicated than
that suited for classic power analysis (Gerrodette 1987; P. C. Johnson
et al. 2015; Bolker 2008). Our simulations support past work that longer
time series are needed when the trend strength (i.e.~rate of increase or
decrease) is weak or when population variability is high (Gerrodette
1987).

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@
